Ages of Planets and Stars Observed by Kepler

### 2024-04-01 paper/ms.tex
update ms

\title{Gyrochronal and Lithium Ages for Stars and Planets Observed by Kepler}

@@

\author{Authors to include (ORDER AFTER LGB NOT YET ASSIGNED MEANING)}

\author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
\affiliation{\caltech}

\author{Lynne~A.~Hillenbrand}
\affiliation{\caltech}

%
% HIRES Collaborators
%
\author[0000-0002-0531-1073]{H. Isaacson}
\affiliation{\berkeley}
%
\author[0000-0001-8638-0320]{A. W. Howard}
\affiliation{\caltech}
%
\author{HIRES observers to be named}
\affiliation{Various}

\begin{abstract}
  We report rotation- and lithium-based ages for stars and planets observed by
  the primary Kepler mission.  Stellar rotation periods are collected from
  previous analyses of the Kepler light curves; lithium content is assessed
  using HIRES spectra.
  We report ages for \nkoisgoldwithages\ FGK stars with transiting planets, with relative statistical
  precisions $\sigma_t / t$ of XX +YY -ZZ\%; a factor of X.X
  +Y.Y -Z.Z more precise than state-of-the-art isochronal ages for the
  same stars.  The results include XXX planets younger than 1\,Gyr,
  and YYY planets younger than 0.5\,Gyr.
  In the base sample of dwarf FGK stars observed by
  Kepler, we find that the intrinsic age distribution is(/not)
  consistent with a uniform distribution between X and Y billion years
  (Gyr).  In the planet sample, we find a consistent(/inconsistent)
  distribution at $N$-sigma.  Analyzing the completeness of the Kepler
  detection pipeline, we interpret the majority of the change to be
  caused by selection effects (or astrophysics).  We also comment on a
  few previously reported evolutionary trends in the size evolution of
  close-in mini-Neptunes (2-4\,\rearth) and super-Earths
  (1-2\,\rearth), a few of which we recover at comparable
  ($\sim$3$\sigma$) significance to previous work, and one of which
  we note for the first time.
\end{abstract}

@@

Discovering a young planet requires solving two problems: find the
planet, and measure the star's age.  Each problem can be
solved in different ways, yielding many possible permutations for
discovery.  In this manuscript, we will focus on the transit method
for planet discovery, and rotation and lithium-based methods for stellar age
measurement.  Rotation-based age-dating, also known as
gyrochronology, will provide the most information for the largest
number of stars.  The lithium-based technique will provide an
important sanity check for the subset of stars younger than a few
hundred million years.

@@

%TODO mention the RV approach?  at least the Hyades etc work...
A wide range of approaches have, over the past ten years, begun to
establish the subfield of young, close-in planet discovery.
Pioneering discoveries with the prime Kepler mission (hereafter,
``Kepler'') focused on stars with known ages---stars in clusters---and
searched them for transiting planets \citep{Meibom_2013}.  Planets
discovered in this way are assumed to have the same age as the host
cluster, which are the standards for the current astrophysical age
scale.  The resulting ages are precise at the $\approx$10\% level.  K2
and TESS have yielded larger numbers of young planets in clusters than
Kepler, because each of these missions has observed a larger fraction
of the sky, and therefore more stars with known ages
\citep[e.g.][]{Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Zhou_2022,Zakhozhay_2022,Wood_2023}.

An important limitation of the ``cluster-first'' approach is that only
$\approx$1\% of stars within the nearest few hundred parsecs can
currently be associated with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
This means that there are many ``young'' 0.05-1\,Gyr stars in the
field that were not part of the selection function of these searches.
Roughly 40 transiting planets are currently known to be younger than
1\,Gyr, with an age precision $t/\sigma_t > 3$ (see
Figure~\ref{fig:rp_period_age}).  However, Kepler discovered
$\approx$4{,}000 planets; and so barring a major discovery bias
against young planets, we would expect a significant portion of the
Kepler planets---perhaps 10\%, or 400 planets---to be younger than
1\,Gyr.  If one is willing to compromise on age precision, those
planets might be identifiable.

@@

Have these young Kepler planets already been found?  If they have, the
most likely studies to have found them would be those by
\citet{Berger_2020b_rpage}, \citet{David_2021} or
\citet{Petigura_2022}.  The Berger and Petigura studies focused on
isochronal age-dating of individual stars, which is most precise in
the context of Kepler for stars that are beginning to evolve off the
main-sequence.  The \citet{David_2021} study leveraged both isochronal
ages from an earlier analogous study
\citep{Fulton_Petigura_2018_cks_vii}, and (in their core analysis) a
relative gyrochronal age-ranking approach that sorted stars based on
whether they rotated faster or slower than a gyrochrone age of
$\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.

Theoretical predictions for planet evolution in the Kepler sample are
most important in the context of mini-Neptunes and super-Earths, since
these are the most abundant planets.  Pertinent predictions include:
{\it a)} that mini-Neptunes should cool and contract, roughly
following a $R_p \propto t^{-0.1}$ scaling \citep{Gupta_2019}; {\it
b)} that large and close-in mini-Neptunes at the edge of the
photoevaporation desert should shrink over the atmospheric loss
timescale, $t_{\rm loss}$ \citep{Owen_Lai_2018}; {\it c)} that the
abundance of 4-8\,R$_\oplus$ planets with masses of 5-30\,M$_\oplus$
should decrease over $t_{\rm loss}$---the upper-envelope of the
mini-Neptune size distribution should corresponding decrease,
e.g.~from $\approx$4.5\,R$_\oplus$ at $\approx$200\,Myr to
$\approx$3.5\,R$_\oplus$ at $\approx$700\,Myr
\citep[e.g.][]{Rogers_2021}; {\it d)} that the super-Earth to
mini-Neptune ratio, $\theta$, should increase over $t_{\rm loss}$
\citep[e.g.][]{Rogers_2021}, with the mini-Neptune abundance
decreasing by at least a factor of two or three, and the super-Earth
abundance correspondingly increasing {\bf BY HOW MUCH?}); {\it e)} that the
upper boundary of the super-Earth population should increase over the
same timescale as $\theta$ is increasing, since more massive
super-Earth cores should be able to retain their envelopes for longer
than less massive cores.

The details of the predictions are flexible, in that they depend on
the distributions of planetary core masses, initial atmospheric mass
fractions, and initial entropies of the atmospheres after disk
dispersal.  For instance, different sources of heating---from the star
\citep{Owen_Wu_2013,Lopez_Fortney_2014,Jin_2014}, planetary interior
\citep{Gupta_2019}, or even giant impacts
\citep{Biersteker_Schlichting_2019}---can produce very different
histories for the size evolution of planets at these early times
\citep[e.g.][]{Owen_2020}.  The generic prediction that the ``upper
envelope'' of the mini-Neptune size distribution should shrink due to
Kelvin-Helmholtz cooling is likely hard to avoid; predictions however
concerning the ``radius valley'' and movement of planets across it are
dependent on the timescales and prevalence of atmospheric mass loss.
In specific scenarios for initial atmospheric accretion, the radius
gap might form without needing to invoke mass loss at all
\citep{Lee_2022}, in which case there could be very little evolution
of the planetary size distribution between 1 and 4\,\rearth\ over the
first gigayear.

Can stellar age measurements provide lines of evidence for or against
the different models?  Neither {\it (a)} nor {\it (b)} has yet had a
convincing---or even suggestive--- detection reported (see
\citealt{Petigura_2022}).  While the apparent over-abundance of large
very young planets ({\it (c)}) has been noted by multiple authors
\citep[e.g.][]{Bouma_2020,Mann_2022}, its presence is almost certainly
exacerbated by a selection effect against sub-Neptune sized planet
detection at $<$100\,Myr \citep{Zhou_2021,Bouma_2022}.  And finally,
{\it (d)} and {\it (e)} have been explored by
\citet{Berger_2020b_rpage} and the twin \citet{David_2021} and
\citet{Sandoval_2021} studies.  Both \citet{Berger_2020b_rpage} and
\citet{Sandoval_2021} reported that $\theta$ increases with time, at a
statistical significance of $\approx$2-3$\sigma$.  For
\citet{Berger_2020b_rpage}, this conclusion was reached by comparing
the sizes of 85 planets with median isochronal ages younger than
1\,Gyr with a property-matched sample of planets older than 1\,Gyr.
In \citet{Sandoval_2021}, the conclusion was reached by computing the
ratio while performing a Monte Carlo re-sampling procedure from the
isochronal ages and planetary sizes computed by
\citet{Fulton_Petigura_2018_cks_vii}.  Finally, effect {\it (e)} was
reported by \citet{David_2021}, who found that the average location of
the radius valley shifted from $\approx$1.8\,R$_\oplus$ to
$\approx$2.0\,R$_\oplus$ between 1.8 and 3.2\,Gyr.

The fact that stellar ages---and especially isochronal ages---are
correlated with stellar masses and metallicities is a source of stress
for many in this subfield.  \citet{Sandoval_2021} for instance noted a
strong trend in $\theta$ as a function of stellar metallicity in
addition to the trend they noted in age, and cautioned that this could
also explain their observations.  Independently, \citet{Petigura_2018}
and \citet{Petigura_2022} have reported trends between small-planet
occurrence and both stellar metallicity and mass; these parameters
seem, at first glance, to be more important predictors of bulk changes
in the planetary population than age.  Given these known correlations,
it would be interesting to consider population models in which
variables such as metallicity, mass, and age were added in succession
\citep[e.g.][]{Thorngren_2021}.  For such an analysis to be effective,
the measurement precision of these different parameters would ideally
be as high as possible.

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 0.5-2.5\,Gyr, and can also provide useful age limits at
earlier times \citep{Bouma_2023}.  The idea of using a star's
spin-down as a clock is quite old
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  Key
for the present analysis is the fact that we can empirically calibrate
rotation-based ages using measured sequences of rotation periods
in open clusters
(\citealt{Bouma_2023}, and references therein).  This approach
ultimately ties the rotation age scale to the cluster age scale, and
is limited in precision by the intrisic scatter in the $P_{\rm rot}$
sequences, and in accuracy by systematic uncertainties in the evolving
spin-down rates, and in the cluster ages themselves.

Lithium-based age-dating includes two qualitatively distinct regimes:
depletion boundary ages derived for M-dwarfs in star clusters, and the
less precise decline-based ages for individual field FGK stars
\citep{Soderblom_2010}.  Here we focus on the latter approach, which
relies on the gradual empirical observation that the Li abundances of
FGK stars decline as a function of time, for reasons different from
the standard convective mixing and burning scenario (CITE, CITE).
Empirically, this decline has recently been documented by
\citet{Jeffries_2023}, who built a model for how the equivalent width
(EW) of the \ion{Li}{1} 6708\,\AA\ doublet evolves as a function of
stellar effective temperature and age for a set of 6{,}200 stars in 52
open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
stars between $\approx$0.03-0.5\,Gyr, with a strong dependence on
spectral type since the K-dwarfs lose their surface lithium much
faster than G-dwarfs.  The precision of the ages reported by the
\citet{Jeffries_2023} method in this regime are in the range of
0.3-0.5~dex.

The methods by which we select the stellar and planet samples are 
discussed in Section~\ref{sec:selection}.
The age-dating methods are described and validated in
Section~\ref{sec:agemethod}.
The population-level trends for both the parent stellar sample and
the planet sample are discussed in Section~\ref{sec:results}.
A few conclusions are offered in Section~\ref{sec:conclusions}.

@@

This work is focused on stars surveyed by Kepler for which ages can be
inferred using either stellar rotation, lithium, or both.  These
``age-dateable stars'', $\mathcal{S}$, are a relatively small subset
of the \nkeplerstars\ Kepler targets.  Rotation periods provide
the majority of the information, since they have been reported for
$\approx$55{,}000 Kepler stars
\citep[e.g.][]{McQuillan_2014,Santos_2021}.  Measurements of the
\ion{Li}{1} 6708\,\AA\ doublet require high-resolution
($R\gtrsim$10{,}000) spectra, and are only available for a few
thousand Kepler objects of interest (KOIs).  We therefore choose to
define our selection function as Kepler stars with measured
rotation periods ($\mathcal{S}$; Section~\ref{subsec:starsel}), and
Kepler planets whose host stars have measured rotation periods
($\mathcal{P}$; Section~\ref{subsec:planetsel}).

@@

We select stars with rotation periods as determined by previous
investigators.  Studies of stellar rotation across the entire Kepler field
include those by \citet{McQuillan_2014} (\citetalias{McQuillan_2014}),
\citet{Reinhold_2015},
\citet{Santos_2019} (\citetalias{Santos_2019}), and
\citet{Santos_2021} (\citetalias{Santos_2021}).
\citetalias{McQuillan_2014} used an approach
based on the autocorrelation function to detect 34{,}030 rotation
periods for main-sequence Kepler targets cooler than 6500\,K, and excluded
known eclipsing binaries and KOIs.  
\citet{Reinhold_2015} used an iterative Lomb-Scargle based approach to
analyze $\approx$40{,}000 stars with a variability range $R_{\rm
	var}>0.3$\% that were not known EBs, planet candidates, or pulsators,
and reported primary rotation periods for 24{,}124 of these stars.
\citet{Reinhold_2015} also reported secondary periods for two thirds
of the stars with primary periods, which could be caused by
differential rotation or finite spot lifetimes.  
Finally,
\citetalias{Santos_2019} and
\citetalias{Santos_2021} combined a wavelet analysis and
autocorrelation-based approach, and cumulatively reported rotation
periods for 55{,}232 main-sequence and subgiant FGKM stars.
\citetalias{Santos_2019} and \citetalias{Santos_2021} included known
KOIs and binaries, and assigned them specific quality flags.  
The rotation periods of KOIs have received considerable additional
scrutiny
\citep[e.g.][]{Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021}.

%After reviewing the relevant literature and considering the nature of
%the rotation period detection problem,
We chose to adopt the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} as our default rotation periods.
\citetalias{Santos_2021} provides a detailed comparison against
\citetalias{McQuillan_2014}; the brief summary is that the periods
agree for 99.0\% of the 31{,}038 period detections in common between
the two studies.  \citetalias{Santos_2021} classified the 2{,}992
remaining stars from \citetalias{McQuillan_2014} as ``non-periodic''
based on updated knowledge of contaminants (e.g.~giant stars and
eclipsing binaries) and visual inspection.  In addition,
\citetalias{Santos_2021} report rotation periods for 24{,}182
main-sequence and subgiant FGKM stars that were not reported as
periodic by \citetalias{McQuillan_2014}.  Our interpretation is that
these detections are likely more challenging than those from
\citetalias{McQuillan_2014}, perhaps due to a lower variability
amplitude, or a longer rotation period.  A relevant flag for stars
missing \citetalias{McQuillan_2014} periods is included in
Table~\ref{tab:stars}.

@@

As a sanity check on the statistical uncertainties of these rotation
periods, we compared our adopted \citetalias{Santos_2019} and
\citetalias{Santos_2021} periods with those reported by
\citet{McQuillan_2014} and \citet{Mazeh_2015}.
We found that for $P_{\rm rot}\lesssim$15\,days, the two 
datasets at a precision of $\lesssim$0.01$P_{\rm rot}$.  At longer periods of 
$P_{\rm rot}\approx$30\,days, the agreement was typically at the
$\lesssim$0.03$P_{\rm rot}$ level, and
the upper envelope of the period difference tended to increase
linearly with period.
Based on this comparison, we adopted a simple prescription for the
period uncertainties,
such that there are 1\% relative uncertainties below $P_{\rm
rot}=15$\,days, and a linear increase thereafter, with slope set to
guarantee 3\% $P_{\rm rot}$ uncertainties at 30 days.

\subsection{Lithium sample}
A final subcomponent of our analysis involves an age assessment based on
the equivalent width of the \ion{Li}{1} 6708\,\AA\ doublet.  For this
component of the work, we primarily rely on spectra acquired with the
High Resolution Echelle Spectrometer (HIRES;
\citealt{vogt_hires_1994}) on the Keck I 10m telescope.
The majority of these spectra are archival, and were acquired
as part of the California Keck Survey (CKS).
A subset were acquired by our team in our own observing programs.
%TODO quantify.
%TODO compare to berger2018?

@@

\section{Stellar Properties}

@@

>>>>

%\title{Gyrochrone and Lithium Ages for Stars and Planets Observed by Kepler}
\title{Ages of Planets and Stars Observed by Kepler}

@@

\author{Authors to include (order after LGB not assigned meaning;
currently alphabetical)}

% Early experiments for Kepler rotation samples
\author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
\affiliation{\caltech}

% ?
%\author{Lynne~A.~Hillenbrand}
%\affiliation{\caltech}

% HIRES observations and reductions
\author[0000-0002-0531-1073]{H. Isaacson}
\affiliation{\berkeley}
% HIRES observations, reductions, JUMP infrastructure
\author[0000-0001-8638-0320]{A. W. Howard}
\affiliation{\caltech}
%
\author{HIRES observers to be named}
\affiliation{Various}

% <250 words
\begin{abstract}
  We report empirically calibrated rotation- and lithium-based ages
  for planets and stars observed by the primary Kepler mission.
  %
  Stellar rotation periods are collected from previous analyses of
  Kepler data; lithium content is assessed using both new and
  archival HIRES spectra.
  %
  We report ages for \nkoisgoldwithages\ FGK stars with transiting
  planets, with relative statistical precisions $\sigma_t / t$ of XX
  +YY -ZZ\%.  This is a factor of X.X +Y.Y -Z.Z more statistically precise than
  isochronal ages for the same stars, and $\approx$X.X times less
  precise than cluster-based ages.
  %
  The results include XXX planets younger than 1\,Gyr (at 3$\sigma$),
  and YYY planets younger than 0.5\,Gyr (at 3$\sigma$).
  %
  We also report rotation-based ages for XX,XXX stars observed
  by Kepler, aka.~the ``target star sample''.
  %
  In the latter sample of dwarf FGK stars observed by Kepler, we find
  that the gyrochrone age distribution rises linearly from 0 to
  3\,Gyr, with stars 2\,Gyr old being twice as common in the Kepler
  field as stars 1\,Gyr old.
  %
  This qualitatively agrees with previous analyses of the Milky Way's star
  formation history, though it offers greater sensitivity in the range of
  $\approx$0.1-1\,Gyr.
  %
%  The sample of detected Kepler planets has a similar age
%  distribution, though it is slightly deficient at $\lesssim$0.2\,Gyr,
%  likely due to decreased detection sensitivity.
  %
  Combining our stellar age catalog with Kepler's completeness products, we perform
  a hierarchical bayesian analysis to infer the average number of planets per star
  as a function of both stellar and planetary parameters.
  %
  While we recover many previously known correlations with stellar mass and metallicity,
  our analysis also highlights the importance of age in shaping the ``sub-Saturn desert'',
  and in influencing the evolution of the mini-Neptune and super-Earth populations.
  %TODO will this be the case?
%  We also comment on a few previously reported evolutionary trends in
%  the size evolution of close-in mini-Neptunes (2-4\,\rearth) and
%  super-Earths (1-2\,\rearth), a few of which we recover at comparable
%  ($\sim$3$\sigma$) significance to previous work, and one of which we
%  note for the first time.
\end{abstract}

@@

Discovering a young planet requires solving two problems: find the
planet, and measure the star's age.  Each problem can be solved in
various ways, and so there are many paths for discovery.  In this
manuscript, we will focus on planets discovered using transits, and
stellar ages measured based using empirical rotation and lithium-based
methods.  Rotation-based age-dating will provide the most information
for the largest number of stars.  The lithium-based technique will
provide useful information for the subset of stars younger than a few
hundred million years.

@@

%TODO mention the RV approach?  at least the Hyades etc work...
A wide range of approaches have, over the past ten years, begun to
establish the subfield of young, close-in planet discovery.
Pioneering discoveries with the prime Kepler mission (hereafter,
``Kepler'') focused on stars with known ages---stars in clusters---and
searched them for transiting planets \citep{Meibom_2013}.  Planets
discovered in this way are assumed to have the same age as the host
cluster.  The
resulting ages are precise at the $\approx$10\% level.  K2
and TESS have yielded larger numbers of young planets in clusters than
Kepler, because each of these missions has observed a larger fraction
of the sky, and therefore more stars with known ages
\citep[e.g.][]{Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Zhou_2022,Zakhozhay_2022,Wood_2023}.

A limitation of the ``cluster-first'' approach is that only
$\approx$1\% of stars within the nearest few hundred parsecs can
currently be associated with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
This means that there are many ``young'' (0.05-1\,Gyr) stars in the
field that are omitted by cluster-only searches.
Of order 40 transiting planets are currently known to be younger than
1\,Gyr, requiring an age precision $t/\sigma_t > 3$ (see
Figure~\ref{fig:rp_period_age}).  However, Kepler discovered
$\approx$4{,}000 planets; and so unless there is a major discovery
bias, we would expect a larger fraction of the Kepler
planets---perhaps 10\%, or 400 planets---to be younger than 1\,Gyr.

@@

Have these young Kepler planets already been found?  If they have, the
most likely studies to have found them would be those by
\citet{Berger_2020b_rpage}, \citet{David_2021} or
\citet{Petigura_2022}.  The \citeauthor{Berger_2020b_rpage} and
\citeauthor{Petigura_2022} studies focused on
isochronal age-dating of individual stars, which is most precise 
for stars whose luminosities and temperature separate them from the
main-sequence.  The \citeauthor{David_2021} study leveraged both isochronal
ages from an earlier analogous study
\citep{Fulton_Petigura_2018_cks_vii}, and, in their core analysis, a
relative gyrochrone age-ranking approach that sorted stars based on
whether they rotated faster or slower than a gyrochrone age of
$\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.

Theoretical predictions for planet evolution in the Kepler sample are
most important in the context of mini-Neptunes and super-Earths, since
these are the most abundant planets.  Pertinent predictions include:
{\it a)} that mini-Neptunes should cool and contract, roughly
following a $R_p \propto t^{-0.1}$ scaling \citep{Gupta_2019}; {\it
b)} that large and close-in mini-Neptunes at the edge of the
photoevaporation desert should shrink over the atmospheric loss
timescale, $t_{\rm loss}$ \citep{Owen_Lai_2018}; {\it c)} that the
abundance of 4-8\,R$_\oplus$ planets with masses of 5-30\,M$_\oplus$
should decrease over $t_{\rm loss}$---the upper-envelope of the
mini-Neptune size distribution should corresponding decrease,
e.g.~from $\approx$4.5\,R$_\oplus$ at $\approx$200\,Myr to
$\approx$3.5\,R$_\oplus$ at $\approx$700\,Myr
\citep[e.g.][]{Rogers_2021}; {\it d)} that the super-Earth to
mini-Neptune ratio, $\theta$, should increase over $t_{\rm loss}$
\citep[e.g.][]{Rogers_2021}, with the mini-Neptune abundance
decreasing by at least a factor of two or three, and the super-Earth
abundance increasing in tandem; {\it e)} that the
upper boundary of the super-Earth population should increase over the
same timescale as $\theta$ is increasing, since more massive
super-Earth cores should be able to retain their envelopes for longer
than less massive cores.

The details of the predictions are flexible, in that they depend on
the distributions of planetary core masses, initial atmospheric mass
fractions, and initial entropies of the atmospheres after disk
dispersal.  For instance, different sources of heating---from the star
\citep{Owen_Wu_2013,Lopez_Fortney_2014,Jin_2014}, planetary interior
\citep{Gupta_2019}, or even giant impacts
\citep{Biersteker_Schlichting_2019}---can produce very different
histories for the size evolution of planets at these early times
\citep[e.g.][]{Owen_2020}.  Different internal compositions also alter
the predicted size evolution \citep[e.g.][]{2024NatAs.tmp...33B}.  The
generic prediction that the ``upper envelope'' of the mini-Neptune
size distribution should shrink due to Kelvin-Helmholtz cooling is
hard to avoid; predictions however concerning the ``radius valley''
and movement of planets across it are dependent on the timescales and
prevalence of atmospheric mass loss.  In specific scenarios for
initial atmospheric accretion \citep{Lee_2022} and early atmospheric
evolution \citep{2024MNRAS.529.2716R} the radius gap might form
without needing to invoke post-disk dispersal mass loss at all.  In
such scenarios, there could be very little evolution of the planetary
size distribution between 1 and 4\,\rearth\ over the first gigayear.

Can stellar age measurements provide evidence for or against
these different models?  Neither {\it (a)} nor {\it (b)} has yet had a
convincing---or even suggestive--- detection reported (see
\citealt{Petigura_2022}).  While an apparent over-abundance of large
very young planets ({\it (c)}) has been noted by multiple authors
\citep[e.g.][]{2017AJ....153...64M,Bouma_2020}, its presence is almost certainly
exacerbated by a selection effect against sub-Neptune sized planet
detection at $<$100\,Myr \citep{Zhou_2021,Bouma_2022}.  Recent studies have
nonetheless found that when accounting for completeness, there is
$\approx$3$\sigma$ evidence that hot mini-Neptunes are more prevalent in
$\approx$100-500\,Myr old K2 \citep{2023AJ....166..248C} and TESS
\citep{2024arXiv240303261V} stars than in the older Kepler sample
\citep[e.g.][]{2020AJ....159..248K}.
%TODO: more kepler citations...

Regarding the observed mini-Neptune to super-Earth ratio, {\it (d)} and {\it
(e)} were explored by \citet{Berger_2020b_rpage} and the twin
\citet{David_2021} and \citet{Sandoval_2021} studies.  Both
\citet{Berger_2020b_rpage} and \citet{Sandoval_2021} reported that $\theta$
increases with time, at a statistical significance of $\approx$2-3$\sigma$.
For \citet{Berger_2020b_rpage}, this conclusion was reached by comparing the
sizes of 85 planets with median isochronal ages younger than 1\,Gyr with a
property-matched sample of planets older than 1\,Gyr.  In
\citet{Sandoval_2021}, the conclusion was reached by computing the ratio while
performing a Monte Carlo re-sampling procedure from the isochronal ages and
planetary sizes computed by \citet{Fulton_Petigura_2018_cks_vii}.  Finally,
effect {\it (e)} was reported by \citet{David_2021}, who found that the average
location of the radius valley shifted from $\approx$1.8\,R$_\oplus$ to
$\approx$2.0\,R$_\oplus$ between 1.8 and 3.2\,Gyr.

The fact that stellar ages---and especially isochronal ages---are
correlated with stellar masses and metallicities is a source of stress
for many in this subfield.  \citet{Sandoval_2021} for instance noted a
strong trend in $\theta$ as a function of stellar metallicity in
addition to the trend they noted in age, and cautioned that this could
also explain their observations.  Independently, \citet{Petigura_2018}
and \citet{Petigura_2022} have reported trends between small-planet
occurrence and both stellar metallicity and mass; these parameters
seem, at first glance, to be more important predictors of bulk changes
in the planetary population than age.  Given these known correlations,
it can be useful to consider population models in which variables such
as metallicity, mass, and age are added in succession
\citep[e.g.][]{Thorngren_2021}.  An alternative, and perhaps more
rigorous approach, is to jointly infer the number of planets per star
as a function of both planetary and stellar parameters, the latter
including age \citep[e.g.][]{2023AJ....166..209M}.  Such analyses are
most effective when the measurement precision of these different
parameters is as high as possible.

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 0.5-4\,Gyr, and can also provide useful age limits at
earlier times \citep{Bouma_2023}.  The idea of using a star's
spin-down as a clock is quite old
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  Key
for our analysis is the fact that we can empirically calibrate
rotation-based ages using measured sequences of rotation periods
in open clusters
(\citealt{Bouma_2023}, and references therein).  This approach
ultimately ties the rotation age scale to the cluster age scale, and
is limited in precision by the intrisic scatter in the $P_{\rm rot}$
sequences, and in accuracy by systematic uncertainties in the evolving
spin-down rates, and in the cluster ages themselves.
The oldest cluster for which this method has been calibrated is currently
M67 \citep[$\approx$4\,Gyr][]{2022ApJ...938..118D}.

Lithium-based age-dating includes two qualitatively distinct regimes:
depletion boundary ages derived for M-dwarfs in star clusters, and the
less precise decline-based ages for individual field FGK stars
\citep{Soderblom_2010}.  Here we focus on the latter approach, which
relies on the observation that the Li abundances of
partially-convective stars gradually decline as they age
\citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
this decline is debated
\citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}
Empirical understanding has however recently improved due to the work
by \citet{Jeffries_2023}, who built a model for how the equivalent
width (EW) of the \ion{Li}{1} 6708\,\AA\ doublet evolves as a function
of stellar effective temperature and age for a set of 6{,}200 stars in
52 open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
stars between $\approx$0.03-0.5\,Gyr, with a strong dependence on
spectral type since the K-dwarfs lose their surface lithium much
faster than G-dwarfs.  The precision of the ages reported by the
\citet{Jeffries_2023} method in this regime are in the range of
0.3-0.5~dex.

The methods by which we select the stellar and planet samples are 
discussed in Section~\ref{sec:selection}.
We summarize the origin of stellar parameters other than the ages in
Section~\ref{sec:stellarprops}, 
and describe and validate our age-dating methods in
Section~\ref{sec:agemethod}.
The population-level trends for both the parent stellar sample and
the planet sample are discussed in Section~\ref{sec:results}.
A few conclusions are offered in Section~\ref{sec:conclusions}.

@@

This work focuses on stars surveyed by Kepler for which ages can be
inferred using either stellar rotation, lithium, or both.  These
``age-dateable stars'' are a relatively small subset of the
\nkeplerstars\ Kepler targets.  Rotation periods provide the majority
of the information, since they have been reported for
$\approx$55{,}000 Kepler stars
\citep[e.g.][]{McQuillan_2014,Santos_2021}.  
High-resolution measurements of the \ion{Li}{1} 6708\,\AA\ doublet are
available for only a few thousand Kepler objects of interest (KOIs).
We therefore define our selection function as Kepler stars
with measured rotation periods ($\mathcal{S}$;
Section~\ref{subsec:starsel}), and Kepler planets whose host stars
have measured rotation periods ($\mathcal{P}$;
Section~\ref{subsec:planetsel}).

@@

We select stars with rotation periods as determined by previous
investigators.  Studies of stellar rotation across the entire Kepler field
include those by \citet{McQuillan_2014} (\citetalias{McQuillan_2014}),
\citet{Reinhold_2015},
\citet{Santos_2019} (\citetalias{Santos_2019}), and
\citet{Santos_2021} (\citetalias{Santos_2021}).
\citetalias{McQuillan_2014} used an approach
based on the autocorrelation function to detect 34{,}030 rotation
periods for main-sequence Kepler targets cooler than 6500\,K, and excluded
known eclipsing binaries and KOIs.  
\citet{Reinhold_2015} used an iterative Lomb-Scargle based approach to
analyze $\approx$40{,}000 stars with a variability range $R_{\rm
	var}>0.3$\% that were not known EBs, planet candidates, or
  pulsators;
they reported primary rotation periods for 24{,}124 of these stars.
\citet{Reinhold_2015} also reported secondary periods for two thirds
of the stars with primary periods, which could be caused by
differential rotation or finite spot lifetimes.  
Finally,
\citetalias{Santos_2019} and
\citetalias{Santos_2021} combined a wavelet analysis and
autocorrelation-based approach, and cumulatively reported rotation
periods for 55{,}232 main-sequence and subgiant FGKM stars.
\citetalias{Santos_2019} and \citetalias{Santos_2021} included known
KOIs and binaries, and assigned them specific quality flags.  
The rotation periods of KOIs have received considerable additional
scrutiny
\citep[e.g.][]{Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021}.

We choose to adopt the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} as our default rotation periods.
\citetalias{Santos_2021} provides a detailed comparison against
\citetalias{McQuillan_2014}; the brief summary is that the periods
agree for 99.0\% of the 31{,}038 period detections in common between
the two studies.  \citetalias{Santos_2021} classified the 2{,}992
remaining stars from \citetalias{McQuillan_2014} as ``non-periodic''
based on updated knowledge of contaminants (e.g.~giant stars and
eclipsing binaries) and visual inspection.  In addition,
\citetalias{Santos_2021} report rotation periods for 24{,}182
main-sequence and subgiant FGKM stars that were not reported as
periodic by \citetalias{McQuillan_2014}.  Our speculation is that
these reported detections are more challenging than those from
\citetalias{McQuillan_2014}, perhaps due to a lower variability
amplitude, or a longer rotation period.  A relevant flag for stars
missing \citetalias{McQuillan_2014} periods is included in
Table~\ref{tab:stars}.
We verified that the main conclusions of our analysis do not change if
we were to adopt only the periods from \citetalias{McQuillan_2014}.

@@

As a sanity check on the statistical uncertainties of these rotation
periods, we compared our adopted \citetalias{Santos_2019} and
\citetalias{Santos_2021} periods with those reported by
\citet{McQuillan_2014} and \citet{Mazeh_2015}.
We found that for $P_{\rm rot}\lesssim$15\,days, the two 
datasets agree at a precision of $\lesssim$0.01$P_{\rm rot}$.  At longer periods of 
$P_{\rm rot}\approx$30\,days, the agreement was typically at the
$\lesssim$0.03$P_{\rm rot}$ level, and
the upper envelope of the period difference tended to increase
linearly with period.
Based on this comparison, we adopted a simple prescription for the
period uncertainties,
such that there are 1\% relative uncertainties below $P_{\rm
rot}=15$\,days, and a linear increase thereafter, with slope set to
guarantee 3\% $P_{\rm rot}$ uncertainties at 30 days.

\subsection{Lithium sample}
A final subcomponent of our analysis involves an age assessment based on
the equivalent width of the \ion{Li}{1} 6708\,\AA\ doublet.  For this
component of the work, we primarily rely on spectra acquired with the
High Resolution Echelle Spectrometer (HIRES;
\citealt{vogt_hires_1994}) on the Keck I 10m telescope.
Most of these spectra were acquired between 2010 and 2020
as part of the California Keck Survey (CKS).
A subset were acquired by our team in our own observing programs.
%TODO quantify.
%TODO compare to berger2018?

@@

\section{Stellar Properties}
\label{sec:stellarprops}

@@

Also Mathur2023, which does ``magneto-gyrochronology'', including the
Sph indicator in a model of the ages.
